DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-06-22 manuscript/manuscript.tex
various edits; removing tex file of figures that is no longer necessary

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4914}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}

@@

Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C. Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its latest stable version -- v0.6.4, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Pennell et al., 2014),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

\begin{minipage}[t]{17.3cm}
\begin{center}
\includegraphics{../figures/figure1/figure1-new.pdf}
\captionof{figure}{Main DateLife workflow. Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package. Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}
\label{fig:workflow}
\end{center}
\end{minipage}

@@

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013). DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.

The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.

@@

At the time of writing of this manuscript
(Jun 21, 2022),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (\url{https://tree.opentreeoflife.org/curator}). As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload and curate new chronograms to OpenTree themselves. Then, trigger an update of their local \texttt{datelife} database to incorporate the new chronograms and have them available for their current analysis.

@@

\emph{\textbf{Choosing a topology.--}}
DateLife requires a tree topology to summarize age data upon.
We recommend that users provide a tree topology as input from the literature, or one of their own making. If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019). These are known as ``induced'' OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

@@

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb, Ackerly, \& Kembel, 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
BLADJ requires a root age estimate. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.

Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
treePL (Stephen A. Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method
(Michael J. Sanderson, 2002).
The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the calibrations as priors on node ages.

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

@@

\texttt{datelife}'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor.
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).

\begin{center}
  \includegraphics{../figures/fig_runtime_main.pdf}
\end{center}
  \captionof{figure}{DateLife's benchmarking results. Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
  \textit{Caption continues.}}

\label{fig:runtime_main}

\begin{justify}
  \textit{Figure caption continued}. 
For each N = \{10, 100, 200, ..., 1 000, ... , 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray). For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).
\end{justify}

\vspace{20mm}

\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat (R Core Team, 2018) that were run both locally
with the devtools package (R Core Team, 2018), and on a public server --via
GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.

@@

\emph{\textbf{Creating a search query.-}}
We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.

@@

\emph{\textbf{Searching the database.-}}
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

\emph{\textbf{Summarizing search results.-}}
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).

\vspace{30mm}

\begin{center}
\includegraphics{../figures/figure2/figure2.pdf}
\end{center}
\vspace{10mm}
\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes. B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names. Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}

@@

\begin{table}[H]
  \caption{Ages of congruified nodes obtained from a datelife search for the Small Example presented. See Figure 3, step C2.}
  \label{tbl:table1}
  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
\end{table}

\begin{table}[H]
  \caption{Summary of congruified nodes ages. See Figure 3, step C3.}
  \label{tbl:table2}
  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
\end{table}

@@

\emph{\textbf{Creating a query.-}}
To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group. Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.

\emph{\textbf{Searching the database.-}}
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).

\emph{\textbf{Summarizing search results.-}}
The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).

@@

We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.

We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).

For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).

@@

\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4924}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced, USA\\\textsuperscript{2} University of Tennessee, Knoxville, USA}

@@

Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its current stable version -- v0.6.5, is available from the The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Pennell et al., 2014),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

\begin{minipage}[t]{17.3cm}
\includegraphics{../figures/figure1/figure1-new.pdf}
\captionof{figure}{\textit{Main DateLife workflow. Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package. Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}}
\label{fig:workflow}
\end{minipage}

@@

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string. Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.

DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013). DateLife implements TNRS with OTT as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing taxonomic identification numbers for further processing.

The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.

@@

At the time of writing of this manuscript
(Jun 22, 2022),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}). As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the \texttt{datelife} package. The updated database is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also implement functions from the \texttt{datelife} R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.

@@

\emph{\textbf{Choosing a topology.--}}
DateLife requires a tree topology to summarize age data upon.
We recommend that users provide a tree topology as input from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

@@

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).

By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology. BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
This minimizes age variance in the resulting chronogram (Webb et al., 2008).
BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
BLADJ requires a root age estimate. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.

Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

@@

\texttt{datelife}'s R package code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor.
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime_main}).

\includegraphics{../figures/fig_runtime_main.pdf}
\captionof{figure}{\textit{DateLife's benchmarking results. Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names. For each N = \{10, 100, 200, ..., 1 000, ... , 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray). For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
\label{fig:runtime_main}

\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat (R Core Team, 2018) that were run both locally
with the devtools package (R Core Team, 2018), and on a public server using the continuous integration tool of GitHub actions (\url{https://docs.github.com/en/actions}).
At present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.

@@

\hypertarget{creating-a-search-query-1}{%
\subsubsection{Creating a search query}\label{creating-a-search-query-1}}

We chose 6 bird species within the Passeriformes. The sample includes
two species of cardinals:
the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
three species of buntings:
the yellowhammer -- \emph{Emberiza citrinella},
the pine bunting -- \emph{Emberiza leucocephalos} and
the yellow-throated bunting -- \emph{Emberiza elegans};
and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.

\hypertarget{searching-the-database}{%
\subsubsection{Searching the database}\label{searching-the-database}}

DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014). No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

@@

\hypertarget{summarizing-search-results-1}{%
\subsubsection{Summarizing search results}\label{summarizing-search-results-1}}

DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).

\begin{center}
\includegraphics{../figures/figure2/figure2.pdf}
\end{center}
\vspace{10mm}
\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes. B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names. Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}

@@

\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
\label{tbl:table1}
\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}

\vspace{15mm}

\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}

\label{tbl:table2}
\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}

@@

\hypertarget{creating-a-query}{%
\subsubsection{Creating a query}\label{creating-a-query}}

To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group. Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.

\hypertarget{searching-the-database-1}{%
\subsubsection{Searching the database}\label{searching-the-database-1}}

Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).

\hypertarget{summarizing-search-results-2}{%
\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}

The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).

@@

We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).

@@

DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.

A total of 99,474 unique terminal taxa are represented in DateLife's database.
Incorporation of more chronograms into the database will continue to improve DateLife's services. One option to increase the number of chronograms in the DateLife database is the Dryad data repository. Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.

\hypertarget{age-variation-in-source-chronograms}{%
\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}
